---
title: "Demonstration of the tuning module of the package"
author: "Shayantan Banerjee"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Demonstration of the tuning module of the package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
options(repos = c(CRAN = "https://cloud.r-project.org/"))

# Suppress title check if necessary
options(rmarkdown.html_vignette.check_title = FALSE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


## Test Dataset for tuning

The test dataset used in this vignette is hosted at [https://zenodo.org/records/15132924](https://zenodo.org/records/15132924). Users need to download the file and unzip the folder contents on their local system.

- The `test_data_MTBLS606` folder contains a folder named `mzml_files/` and two files: `Sample-and-feature-wise-RT-tolerance.txt` and `feature_list.txt`. The `feature_list.txt` file has three columns: "ID", "mz", and "RT". The additional file `Sample-and-feature-wise-RT-tolerance.txt` contains the sample-wise RT tolerances for all features. The number of rows in both files is the same, while the `Sample-and-feature-wise-RT-tolerance.txt` file has twice as many columns as `feature_list.txt`, with `RT_min_` and `RT_max_` appended to every sample name.
  
## Install the dependencies
A simple and automated way to install and load all the dependencies is to run the following function
```{r initial-setup, eval=FALSE}
install_and_load_dures_dependencies()
```
```


## Install the package
```{r initial-setup-contd, eval=TRUE}
if (!requireNamespace("devtools", quietly = TRUE)) {
  install.packages("devtools", quiet = TRUE)
}
devtools::install_github("BiosystemEngineeringLab-IITB/dures")
library(dures)
## First Test Example
## Test Dataset
```

The test dataset can be downloaded from the data deposited at [https://zenodo.org/records/15132924](https://zenodo.org/records/15132924) (folder name: `test_data_MTBLS606/`).

### Workflow Overview

1. **Reading mzML Files**  
   The first step of the workflow involves reading mzML files and extracting MS/MS spectra within the specified *m/z* and *RT* tolerance.

2. **Extracting and Concatenating Spectra**  
   The extracted spectra are then concatenated for all replicate spectra belonging to a given feature.

3. **Filtering Features**  
   Features that do not have any MS/MS spectra extracted, based on the specified *RT* and *m/z* tolerance, are removed from the list.

```{r, execute first step of the workflow}
# Define the path to the  folder (replace with the actual path on your system)
folder_path <- "/data1/sbanerjee/test_data_MTBLS606/"

# Check if the directory exists
if (!dir.exists(folder_path)) {
  stop("The directory test_1 does not exist. Please modify the path to the actual location of the test_1 folder")
}

l1 <- preprocess(folder_path = folder_path, tol_mz = 25, tol_rt = 0.1)
```


More details on what the list `l1` contain, is provided in the other vignette titled: "Demonstration of the package dures"

