---
title: "Demonstration of the package dures"
author: "Shayantan Banerjee"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Demonstration of the package dures}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
options(repos = c(CRAN = "https://cloud.r-project.org/"))

# Suppress title check if necessary
options(rmarkdown.html_vignette.check_title = FALSE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The **Denoising Using Replicate Spectra (DuReS)** package requires the following input parameters:

- A folder path to store all the data and analysis results.
- A folder named `mzml_files/` within the above folder path, containing files in the mzML format.
- A features file containing a list of MS/MS features derived from standard untargeted metabolomics analysis software. This file must satisfy one of the following requirements:
  - **Option 1**: A single file named `feature_list.txt` with columns “ID”, “mz”, and “RT”. In this case, a fixed but customizable RT and mz tolerance will be used to extract the MS/MS spectra.
  - **Option 2**: Both:
    - A `feature_list.txt` file in the format described above, and
    - A RT tolerance file named `Sample-and_feature-wise-RT-tolerance.txt` containing custom RT tolerances for each sample. For example:
      - If there are two samples (`sample_1` and `sample_2`), the `Sample-and_feature-wise-RT-tolerance.txt` file should have the same number of rows as `feature_list.txt`.
      - The column names should be formatted as `RT_min_sample_1`, `RT_min_sample_2`, `RT_max_sample_1`, and `RT_max_sample_2`.


## Test Datasets

All the test datasets used in this vignette are hosted at [https://zenodo.org/records/13778168](https://zenodo.org/records/13778168). Users need to download the file and unzip the folder contents on their local system.

- The `test_1` folder contains a folder named `mzml_files/` and two files: `Sample-and-feature-wise-RT-tolerance.txt` and `feature_list.txt`. The `feature_list.txt` file has three columns: "ID", "mz", and "RT". The additional file `Sample-and-feature-wise-RT-tolerance.txt` contains the sample-wise RT tolerances for all features. The number of rows in both files is the same, while the `Sample-and-feature-wise-RT-tolerance.txt` file has twice as many columns as `feature_list.txt`, with `RT_min_` and `RT_max_` appended to every sample name.
  
- The `test_2` and `test_3` folders each contain a folder named `mzml_files/` and a file named `feature_list.txt`.


## Install the dependencies
```{r initial-setup, eval=FALSE}
# Dependencies
install.packages("BiocManager", quiet = TRUE)
BiocManager::install(c("Spectra", "S4Vectors", "mzR"), quiet = TRUE)

if (!requireNamespace("devtools", quietly = TRUE)) {
  install.packages("devtools", quiet = TRUE)
}
```

## Install the package
```{r initial-setup-contd, eval=TRUE}
devtools::install_github("BiosystemEngineeringLab-IITB/dures")
library(dures)
library(Spectra)
library(mzR)
```

## First Test Example
## Test Dataset

The test dataset can be downloaded from the data deposited at [Zenodo](https://zenodo.org/records/13778168) (folder name: `test_1/`).

### Workflow Overview

1. **Reading mzML Files**  
   The first step of the workflow involves reading mzML files and extracting MS/MS spectra within the specified *m/z* and *RT* tolerance.

2. **Extracting and Concatenating Spectra**  
   The extracted spectra are then concatenated for all replicate spectra belonging to a given feature.

3. **Filtering Features**  
   Features that do not have any MS/MS spectra extracted, based on the specified *RT* and *m/z* tolerance, are removed from the list.

```{r, execute first step of the workflow}
# Define the path to the test_1 folder (replace with the actual path on your system)
folder_path <- "~/Desktop/test_package_dures/test_1/"
# Check if the directory exists
if (!dir.exists(folder_path)) {
  stop("The directory 'test_1' does not exist. Please modify the path to the actual location of the 'test_1' folder.")
}
l1 = preprocess(folder_path = folder_path, tol_mz = 5, tol_rt = 0.1)
```
`l1` is a list containing two elements:

1. **`stats_file_ms2_only`**: This element contains a dataframe with MS/MS feature information, including:
   - **ID**: Feature identifier
   - **mz**: Mass-to-charge ratio
   - **RT**: Retention time
   - **mz_up**: Upper mass-to-charge limit
   - **mz_down**: Lower mass-to-charge limit
   - **RT_tolerance**: Retention time tolerance for each sample

2. **`spectra_ms2_only`**: This element contains a `Spectra` object with the replicate spectra concatenated for each feature.

Now, let us examine the concatenated spectra for the **first feature**.


```{r first element}
cat("name of first feature\n")
names(l1$spectra_ms2_only)[1] #name of the first feature
l1$spectra_ms2_only[[1]]
```
```{r second}
cat("name of third feature\n")
names(l1$spectra_ms2_only)[3] #name of the first feature
l1$spectra_ms2_only[[3]]
```
From the output, we can see that the first two MS2 features (`1982` and `872`) have **83** and **43** MS/MS spectra, respectively, from different samples listed in the `file(s)` section of the output.

We can quickly take a look at the peak information of the first two out of the 83 raw MS/MS spectra using the `peaksData()` command. Please ensure that the **Spectra** package is loaded in your workspace.


```{r peaksdata}
peaksData(l1$spectra_ms2_only[[1]]$aggregate)[[1]] #first replicate spectrum
peaksData(l1$spectra_ms2_only[[2]]$aggregate)[[1]] #second replicate spectrum
```

### Tracking Individual Spectrum

It is also worth noting that we keep track of each individual spectrum with respect to the sample from which it is derived. Below, we demonstrate this with the identities of the **first 10 replicate spectra**:

```{r spectrum identity}
l1$spectra_ms2_only[[1]]$aggregate$spectrumId[1:10] #will be a vector of length 83
```
### Demonstration of Spectrum Analysis

For demonstration purposes, we will consider the **first spectrum** from the list, named:

> **`20092020_CovidMild_P291_P_2.mzML_scan_1504`**

This spectrum is derived from the sample:

> **`20092020_CovidMild_P291_P_2.mzML`**

As seen in the previous code block, this spectrum contains **19 fragments**.

Similarly, the **second spectrum**:

> **`20092020_CovidMild_P291_P_2.mzML_scan_1541`**

contains **46 fragments**.

At the end of this demonstration, we will see how many fragments remain in these two spectra after applying a **fixed frequency cutoff** to illustrate the effect of **denoising**.

---

### Step 2: Extracting Top x% TIC Spectra and Grouping Fragments

In the second step, we will use `l1` from the previous step as input to:

1. **Extract the top x% TIC spectra**, and
2. **Group fragments** within a specified **mass tolerance**.

The user can modify the following parameters:
- **Top x% TIC value**: Default is **80%** (i.e., **0.8**).
- **Mass tolerance**: Default is **0.05 Da**.

---


```{r second step}
l2 = extract_raw_spectra(folder_path = folder_path, l1, 0.05, 0.8)
```
The following code checks to ensure that the object `sps_top80_tic_2` is available in the local environment. This is not required while running the package

```{r check_and_assign, echo=TRUE, error=TRUE}
# Check if 'sps_top80_tic_2' exists before assigning
if (!exists("sps_top80_tic_2")) {
  stop("Object 'sps_top80_tic_2' not found in the environment.")
} else {
  assign("sps_top80_tic_2", sps_top80_tic_2, envir = .dures_env)
}
```
### Step 2 Output Overview

For the same number of MS/MS features present in `l1`, the output from the second step (`l2`) contains three components:

1. **Top x% TIC Spectra**:  
   Stored as `sps_top_tic_2` in **Spectra** format.
  
2. **Dataframe** (`df`):  
   Contains the number of spectra **before** and **after** applying the top x% TIC cutoff.

3. **Vector of Features without Extracted Spectra**:  
   Contains the feature IDs for which no spectra could be extracted.

### Exploring the Data

#### 1. Viewing Specific Rows of Dataframe `df`

Let's look at the **first** and **third rows** of `df`and the number of spectra after applying top x% TIC cutoff:


```{r intra}
l2$df[c(1,3),]
print(l2$sps_top_tic_2[[1]]) # From 83 before to 66 spectra after after top x% TIC
print(l2$sps_top_tic_2[[3]]) # From 43 before to 34 spectra remain after top x% TIC
```

### Exploring the Spectra Sizes

We will examine the **size** of the **first** and **third spectra** belonging to feature IDs `1982` and `872`, respectively.

#### Intra-Spectrum Grouping

- We performed **intra-spectrum grouping**, meaning that fragments within a default tolerance of **0.05 Da** of one another were merged:
  - **Mean of m/z** values.
  - **Sum of intensities**.

- The resulting spectra are stored in the folder `MS2_scans_before_denoising`.

- Due to intra-spectrum grouping, we can clearly see a **reduction in the number of peaks**.

### Ensuring Consistency Before Comparison

Before comparing the spectra **before and after Step 2** of the analysis, we need to check their **identity** to ensure we are comparing the **same spectra**.

---

```{r compare_spectra}
print(l1$spectra_ms2_only[[1]]$aggregate$spectrumId[1:10]) #spectra identities before step 2
print(l2$sps_top_tic_2[[1]]$spectrumId[1:10]) #spectra identitied after x% TIC cutoff implementation in step 2
```
### Interpretation of Step 2 Output

It is clear from this output that some spectra have been eliminated due to low TIC. 

- **Example**:  
  The spectrum titled `20092020_CovidMild_P291_P_2.mzML_scan_1504` **did not exist** after applying the top 80% TIC cutoff.

#### Exploring Feature 1982

- For the **first feature**, `1982`, we will consider the spectrum titled `20092020_CovidMild_P294_P_2.mzML_scan_1587` (which is the **third on the list** above).

- Let’s look at the **number of fragments before Step 2**:

```{r peaksdata_before_intra}
peaksData(l2$sps_top_tic_2[[1]][3])[[1]]
```
### Intra-Spectrum Grouping Overview

There are **23 fragments** before intra-spectrum grouping. Now, if we look at the same spectrum from the folder `MS2_scans_before_denoising`:

- **Subfolder Name**: `1982`
- **Text File Name**: `20092020_CovidMild_P294_P_2.mzML_scan_1587.txt`

After intra-spectrum grouping, there are **22 fragments**.

The fragments with m/z values **304.11557** and **304.15106** were combined into a single fragment after the intra-spectrum grouping process.


```{r after_grouping}
spectrum_size_after_grouping = read.delim("~/Desktop/test_package_dures/test_1/MS2_scans_before_denoising/1982/20092020_CovidMild_P294_P_2.mzML_scan_1587.txt")
print(dim(spectrum_size_after_grouping)[1])
```

### Another example to demonstrate the effect of intra-spectrum grouping

Similar observations were made for the metabolite ID `872` using the spectrum file:

**`13092020_CovidMild_P282_P_2.mzML_scan_1120`**

The analysis included:

- **Original Number of Fragments**:
  - Before grouping: `98` fragments.
  
- **Corresponding Entry Location**:
  - The original fragments can be found in the folder: `MS2_scans_before_denoising`.

- **Fragments After Grouping**:
  - After grouping, the number of fragments reduced to `81`.

This significant reduction from **98** to **81** fragments demonstrates the effectiveness of the grouping process.


```{r another example}
print(head(peaksData(l2$sps_top_tic_2[[3]][1])[[1]]))
print(dim(peaksData(l2$sps_top_tic_2[[3]][1])[[1]])[1])
```
```{r after_grouping_1}
spectrum_size_after_grouping = read.delim("~/Desktop/test_package_dures/test_1/MS2_scans_before_denoising/872/13092020_CovidMild_P282_P_2.mzML_scan_1120.txt")
print(dim(spectrum_size_after_grouping)[1])
```
### Step 3: Consensus Spectrum Generation

In the **third step**, a **consensus spectrum** is generated using the **top 80% TIC spectra**, and the corresponding **fragment frequencies** are calculated.

- The **consensus spectrum**, also known as the **inter-spectrum aggregate**, is built using the same principle as the **intra-spectrum aggregate**.
- The output from this step includes:
  1. A **single spectrum per feature**.
  2. The **peak frequencies** for each feature.

```{r third step}
l3 = call_aggregate(l2$sps_top_tic_2, 0.05, folder_path)
```

### Examining the Output of `l3`

The output of `l3` contains the following components:

1. **Aggregate Spectra**:  
   Represented as a **Spectra object**.

2. **Dataframe of Aggregate Spectra**:  
   This dataframe includes:
   - **Mean m/z**: The average m/z value of the spectra.
   - **Mean Intensity**: The average intensity of the spectra.
   - **Frequency**: The frequency of occurrence for each fragment.
   - **Number of Fragments**: Before and after **inter- and intra-spectra grouping**.

### Example: Feature `1982`

Let’s consider the first feature with **feature ID `1982`** as an example to examine the output more closely.

```{r inspect-l3}
print(head(l3[[1]]$Df))
print(dim(l3[[1]]$Df)[1])
l3[[1]]$Mean #will return one consensus spectrum

```
The consensus spectrum built using top 80% TIC spectra for feature `1982` has 498 fragments. Every fragment contain the recurrence frequencies. Let's see some of the top recurrent fragments.

```{r recur}
l = l3[[1]]$Df
print(head(l[order(l$Frequency, decreasing = TRUE),],10))
```

### Step 4: Labelling individual spectra for every feature

### Step 4: Labeling Fragments of Top TIC Spectra

Using the **recurrence frequencies** calculated in **Step 3**, we will now label the fragments of every **top TIC spectra** for a given feature.

The output from this step is:

- A **list** containing all spectra for a given feature.
- The corresponding **fragment frequencies** for each labeled fragment.

This labeling process provides insights into how often each fragment appears across all spectra, allowing for better analysis of fragment patterns within the **top TIC** data.


```{r label_spec}
l4 = label_individual_spectrum(l3, folder_path, 0.05)
```

### Inspecting Feature `1982`

We will now inspect the **first feature** (`1982`) and check the corresponding **spectra** and their **fragment frequencies**.

- The **number of spectra per feature** will be the same as in **Step 2**, where we retained only the **top x% TIC spectra**.

- Let's also look at the specific feature named `20092020_CovidMild_P294_P_2.mzML_scan_1587`, as in **Step 2**.


```{r check step4}
print(length(l4[[1]]))
print(names(l4[[1]][3]))
print(l4[[1]][3])
```

### Overview of Fragment Recurrence

We can observe **22 fragments** from the spectrum and their corresponding **frequency of recurrence**.

### Next Steps

In the next step, we will:

1. **Apply a Fixed Frequency Cutoff**:  
   - Denoise the data by retaining only those fragments that meet the cutoff criteria.

2. **Aggregate All MS/MS Spectra**:  
   - Combine all the spectra from the different samples.

3. **Export the Results**:  
   - Print the aggregated MS/MS spectra to a file for further analysis.
   
```{r last_step}
l5 = generate_denoised_spectra(l4, folder_path, ion_mode = "pos") 
```

### Output Overview

The output consists of **mzML format** files with only **denoised MS/MS features**. The output from `l5` contains:

- mzML files saved in a folder named **`Denoised_spectra`**.

### Checking Specific Spectrum Output

We will:

1. **Check the Size and Contents** of a Particular Spectrum
   - Spectrum titled **'1587'**.
   - From the sample **"20092020_CovidMild_P294_P_2"**.
   - Feature ID: **`1982`**.
   
2. **Read the Spectrum from `Denoised_spectra` Folder**: 

`denoised_spectrum <- read_mzML("Denoised_spectra/20092020_CovidMild_P294_P_2.mzML")`


***Number of peaks before denoising***
```{r eg}
#20092020_CovidMild_P294_P_2.mzML_scan_1587
backend = Spectra::MsBackendMzR()
sps <- Spectra::Spectra("~/Desktop/test_package_dures/test_1/Denoised_spectra/20092020_CovidMild_P294_P_2.mzML", source = backend)
print(peaksData(l2$sps_top_tic_2[[1]][3])[[1]])
```

***Number of peaks after intra-spectrum grouping and labelling individual peaks***

```{r test 5}
print(l4[[1]][3])
```

***Number of peaks remaining after applying fixed intensity threshold***

```{r test 6}
print(peaksData(sps)[[1]])
```

### Summary of Results

After applying all **five steps** of the package, only **9 fragments** remained.

The **mzML files** containing the cleaned MS/MS spectra are saved in the folder `Denoised_spectra`.

These files can now be used for further analysis, such as **metabolite annotation**.
