---
title: "Demonstration of the package dures"
author: "Shayantan Banerjee"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The **Denoising Using Replicate Spectra (DuReS)** package requires the following input parameters:

- A folder path to store all the data and analysis results.
- A folder named `mzml_files/` within the above folder path, containing files in the mzML format.
- A features file containing a list of MS/MS features derived from standard untargeted metabolomics analysis software. This file must satisfy one of the following requirements:
  - **Option 1**: A single file named `feature_list.txt` with columns “ID”, “mz”, and “RT”. In this case, a fixed but customizable RT and mz tolerance will be used to extract the MS/MS spectra.
  - **Option 2**: Both:
    - A `feature_list.txt` file in the format described above, and
    - A RT tolerance file named `Sample-and_feature-wise-RT-tolerance.txt` containing custom RT tolerances for each sample. For example:
      - If there are two samples (`sample_1` and `sample_2`), the `Sample-and_feature-wise-RT-tolerance.txt` file should have the same number of rows as `feature_list.txt`.
      - The column names should be formatted as `RT_min_sample_1`, `RT_min_sample_2`, `RT_max_sample_1`, and `RT_max_sample_2`.


## Test Datasets

All the test datasets used in this vignette is hosted at https://zenodo.org/records/13778168. The user needs to download the file and unzip the contents of the folder on their local system. 
- `test_1` folder contains 

## Styles

The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows:

    output: 
      rmarkdown::html_vignette:
        css: mystyles.css

## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))
