---
title: "Demonstration of the tuning module of the package"
author: "Shayantan Banerjee"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Demonstration of the tuning module of the package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
options(repos = c(CRAN = "https://cloud.r-project.org/"))
#this code block will
# Clean your folder_path every time the vignette is run
# 
# Prevent file conflicts and stale output errors
# 
# Ensure reproducibility for every re-render
# Suppress title check if necessary
options(rmarkdown.html_vignette.check_title = FALSE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

# === Custom folder cleanup ===

# Define the base folder path
folder_path <- "/data1/sbanerjee/test_data_MTBLS606/"  # Update this to your actual working path to the test data 

# Files and folders to keep
keep_files <- c("feature_list.txt", "Sample-and-feature-wise-RT-tolerance.txt", "mzml_files")

# List all contents in folder_path
all_contents <- list.files(folder_path, full.names = TRUE)

# Delete everything except essentials
to_delete <- all_contents[!basename(all_contents) %in% keep_files]
unlink(to_delete, recursive = TRUE, force = TRUE)

message("✅ Cleaned folder_path — kept only required files/folders.")
```



## Test Dataset for tuning

The test dataset used in this vignette is hosted at [https://zenodo.org/records/15132924](https://zenodo.org/records/15132924). Users need to download the file and unzip the folder contents on their local system.

- The `test_data_MTBLS606` folder contains a folder named `mzml_files/` and two files: `Sample-and-feature-wise-RT-tolerance.txt` and `feature_list.txt`. The `feature_list.txt` file has three columns: "ID", "mz", and "RT". The additional file `Sample-and-feature-wise-RT-tolerance.txt` contains the sample-wise RT tolerances for all features. The number of rows in both files is the same, while the `Sample-and-feature-wise-RT-tolerance.txt` file has twice as many columns as `feature_list.txt`, with `RT_min_` and `RT_max_` appended to every sample name.
  
## Install the dependencies
A simple and automated way to install and load all the dependencies is to run the following function
```{r initial-setup, eval=FALSE}
install_and_load_dures_dependencies()
```
This should be followed by installing the actual package using devtools as shown below

```{r initial-setup-contd, eval=TRUE}
if (!requireNamespace("devtools", quietly = TRUE)) {
  install.packages("devtools", quiet = TRUE)
}
devtools::install_github("BiosystemEngineeringLab-IITB/dures", auth_token = NULL)
library(dures)
```

## First Test Example
## Test Dataset

The test dataset can be downloaded from the data deposited at [https://zenodo.org/records/15132924](https://zenodo.org/records/15132924) (folder name: `test_data_MTBLS606/`).

### Workflow Overview

1. **Reading mzML Files**  
   The first step of the workflow involves reading mzML files and extracting MS/MS spectra within the specified *m/z* and *RT* tolerance.

2. **Extracting and Concatenating Spectra**  
   The extracted spectra are then concatenated for all replicate spectra belonging to a given feature.

3. **Filtering Features**  
   Features that do not have any MS/MS spectra extracted, based on the specified *RT* and *m/z* tolerance, are removed from the list.

```{r, execute first step of the workflow}
# Define the path to the  folder (replace with the actual path on your system)
folder_path <- "/data1/sbanerjee/test_data_MTBLS606/"

# Check if the directory exists
if (!dir.exists(folder_path)) {
  stop("The directory test_1 does not exist. Please modify the path to the actual location of the test_1 folder")
}

l1 <- preprocess(folder_path = folder_path, tol_mz = 25, tol_rt = 0.1)
```


More details on what the list `l1` contain, is provided in the other vignette titled: "Demonstration of the package dures"

```{r, execute intermediate step of workflow}

print(paste0("Length of stats file after removing features from which MS2 spectra couldn't be extracted: ", as.character(dim(l1$stats_file_ms2_only)[1])))
```
For purposes of this vignette, we will consider a random subset of 50 features from this list of 141 MS2 features. The following step can be skipped while runnning the code on an actual experimental dataset.

```{r, Subsetting the feature list to reduce computation}
# Set seed for reproducibility
set.seed(123)

# Step 1: Get all IDs from the dataframe
all_ids <- l1$stats_file_ms2_only$ID

# Step 2: Sample 50 IDs
selected_ids <- sample(all_ids, 2)

# Step 3: Subset dataframe and reorder by selected_ids
subset_stats <- l1$stats_file_ms2_only[l1$stats_file_ms2_only$ID %in% selected_ids, ]
subset_stats <- subset_stats[match(selected_ids, subset_stats$ID), ]

# Step 4: Subset and reorder spectra list to match the same order
subset_spectra <- l1$spectra_ms2_only[selected_ids]  # list is named by ID

# Step 5: Store both in a new list
l1_subset <- list(
  stats_file_ms2_only = subset_stats,
  spectra_ms2_only = subset_spectra
)
```

From now on, we will use `l1_subset` for all subsequent analysis. The user may stick to `l1` to continue with the original set of MS2 features

### Step 2: Extracting Top x% TIC Spectra and Grouping Fragments

In the second step, we will use `l1` from the previous step as input to:

1. **Extract the top x% TIC spectra**, and
2. **Group fragments** within a specified **mass tolerance**.

The user can modify the following parameters:
- **Top x% TIC value**: Default is **80%** (i.e., **0.8**).
- **Mass tolerance**: Default is **0.05 Da**.

---


```{r second step}
l2 = extract_raw_spectra(folder_path = folder_path, l1_subset, 0.05, 0.8)
```

The following code checks to ensure that the object `sps_top80_tic_2` is available in the local environment. This is not required while running the package

```{r check_and_assign, echo=TRUE, error=TRUE}
# Check if 'sps_top80_tic_2' exists before assigning
if (!exists("sps_top80_tic_2", envir = .dures_env)) {
  stop("Object 'sps_top80_tic_2' not found in .dures_env.")
} else {
  assign("sps_top80_tic_2", get("sps_top80_tic_2", envir = .dures_env), envir = .dures_env)
}

```

Please refer to the vignette titled "Demonstration of the package dures" to know more about the structure of `l2`

### Step 3: Consensus Spectrum Generation

In the **third step**, a **consensus spectrum** is generated using the **top 80% TIC spectra**, and the corresponding **fragment frequencies** are calculated.

- The **consensus spectrum**, also known as the **inter-spectrum aggregate**, is built using the same principle as the **intra-spectrum aggregate**.
- The output from this step includes:
  1. A **single spectrum per feature**.
  2. The **peak frequencies** for each feature.

```{r third step - most expensive}
l3 = call_aggregate(l2$sps_top_tic_2, 0.05, folder_path)
```
### Optional step: Install Git LFS to access curated libraries

The curated reference spectral libraries (`library_positive.rds` and `library_negative.rds`) used for MS/MS matching are tracked using **Git Large File Storage (LFS)** due to their size.

#### 🔧 Step-by-Step Instructions

1. **Install Git LFS (if not already installed)**

   - **Windows/macOS**:  
     Download from [https://git-lfs.github.com](https://git-lfs.github.com)

   - **Linux (Debian/Ubuntu)**:  
     ```bash
     sudo apt install git-lfs
     ```

2. **Initialize Git LFS in your local repo**
   ```bash
   git lfs install
   ```

3. **Clone the repository (if you haven’t already)**  
   ```bash
   git clone https://github.com/BiosystemEngineeringLab-IITB/dures.git
   cd dures
   ```

4. **Pull the large files managed by Git LFS**
   ```bash
   git lfs pull
   ```

5. After pulling, you will find:
   - `library_positive.rds`
   - `library_negative.rds`

   These files are now accessible and can be directly loaded in R using:
   ```r
   library_positive <- readRDS("path/to/library_positive.rds")
   library_negative <- readRDS("path/to/library_negative.rds")
   ```

> **Note:** These libraries are essential for precursor and fragment-level matching in the `dures` workflow.




### Step 4: Labelling individual spectra for every feature

Using the **recurrence frequencies** calculated in **Step 3**, we will now label the fragments of every **top TIC spectra** for a given feature.

The output from this step is:

- A **list** containing all spectra for a given feature.
- The corresponding **fragment frequencies** for each labeled fragment.

This labeling process provides insights into how often each fragment appears across all spectra, allowing for better analysis of fragment patterns within the **top TIC** data. This can be an expensive operation, especially if there are a lor of fragments per feature.


```{r label_spec}
l4 = label_individual_spectrum(l3, folder_path, 0.05)
```
### Step 5: Matching Precursor m/z with Reference Library

In this step, we match the **precursor *m/z*** values of the selected 50 features against a **curated reference library** (included in the package) using a user-defined tolerance.

Based on the specified **ionization mode**, a corresponding set of reference spectra is generated. Additionally, **fragments in the reference spectra** that fall within the specified *m/z* tolerance (`tol`) are grouped together to support more robust **MS/MS-level fragment matching**.

> **Note:**  
> - We are using `l1_subset` (not `l1`) for this analysis.  
> - This step requires the object generated in **Step 2**, namely `sps_top80_tic_2`.  
>   Ensure that **Step 2 has been executed** prior to running this section.

```{r, reference generation}
l5 = precursor_matching(l1_subset, folder_path = folder_path, "negative", 0.05)
```
After this step, a folder named `Reference` will be created inside your specified `folder_path`.  
This folder contains the processed reference spectra that will be used for downstream MS/MS-level matching.

A message will also be displayed listing all precursor ions that **did not match** with any reference spectra within the given tolerance.

Upon inspecting the object l5, you will notice that it is just a modified version of the stats file with only those precursors which yielded a library match. 

```{r inspecting l5}
print(paste0("Number of features removing after this step: ", dim(l5)[1], " out of the original ", dim(l1_subset$stats_file_ms2_only)[1]))
```

### Step 6: Fragment matching before denoising

Since we are currently tuning parameters on the available MS/MS spectra, we begin by matching each spectrum with entries from the `Reference` library. For each spectrum, we identify and record the **reference spectrum entry** that yields the **best match** before denoising.

This step is essential for tracking how well a spectrum matches a known reference in its raw form. After applying denoising using the same experimental spectrum, we will re-match it with **the same reference spectrum** to evaluate changes in similarity scores.

This before/after comparison allows us to quantify the effect of denoising using objective metrics such as similarity score, dot product, or fragment match rate. 
The input to the next function includes:

- The dataframe `l5`
- The MS/MS fragment ion tolerance (`tol`)
- The ionization mode (e.g., `"positive"` or `"negative"`)

In this step, **all available MS/MS spectra** corresponding to each precursor in `l5` are matched against **all available library spectra** from the reference database. For each spectrum, the **best-matching reference spectrum** is recorded based on a chosen similarity metric.

> ⚠️ **Note:**  
> This step is **computationally intensive**, as it performs exhaustive pairwise matching between experimental and reference spectra.


```{r, fragment_match_before_denoising}
l6 = fragment_matching_before_denoising(folder_path, l5, 0.05, "negative")
```

The corresponding results are stored in the folder `Before_denoising_matches`. Let us inspect this. You can change this to an ID of your choice. 

```{r view before denoising results}
id = "ID_868"
file = read.csv(paste0(folder_path, "Before_denoising_matches/", id, ".csv"))
# Display the first few rows and columns (e.g., first 5 rows and 5 columns)
head(file[, 1:5])
# Generate summary statistics for all numeric columns
summary(file)
```
### Description of Columns in the Matching Results (`ID_868.csv`)

The table below summarizes the meaning and role of each column in the MS/MS spectral matching results:

| **Column Name**                  | **Description**                                                                 |
|----------------------------------|---------------------------------------------------------------------------------|
| `Matching_Score`                 | Overall similarity score between the experimental and reference spectra.       |
| `Modified_Dot_Product`          | A variant of the dot product that adjusts for fragment intensity distribution. |
| `Number_of_Matching_Fragments`  | Number of fragment ions that matched between experimental and reference spectra.|
| `Forward_Dot_Product`           | Dot product calculated with the experimental spectrum as the reference frame.  |
| `Reverse_Dot_Product`           | Dot product calculated with the library spectrum as the reference frame.       |
| `Total_Experimental_fragments`  | Total number of fragment ions in the experimental spectrum.                    |
| `Total_Library_fragments`       | Total number of fragment ions in the reference library spectrum.               |
| `Feature_ID`                    | Identifier for the experimental feature or precursor ion.                      |
| `Scan_Number`                   | Identifier for the individual scan associated with the spectrum.               |
| `Library_name`                  | Name or accession of the matched library entry.                                |
| `Splash_key`                    | Unique SPLASH identifier for the library spectrum (if available).              |
| `ID`                            | Internal identifier combining feature ID and scan information.                 |
| `mz`                            | Precursor *m/z* value for the matched spectrum.                                |

These columns are essential for evaluating the quality of spectral matches both before and after denoising. Metrics like `Matching_Score`, `Dot_Products`, and `Number_of_Matching_Fragments` can be compared to assess improvement in spectral quality post-denoising.

### Step 7: Deriving the optinal recurrence frequency cutoff `tuning_module()`

This function performs **frequency-based tuning** of MS/MS spectra denoising parameters for a curated set of precursor features. It evaluates multiple frequency thresholds to identify optimal denoising cutoffs using a Pareto optimization framework.

---

##### **Inputs**

- `folder_path`: Path to the working directory containing input and output folders.
- `l4`: Output from `label_individual_spectrum()` — a list of MS/MS spectra annotated with fragment frequencies.
- `l5`: Data frame of features matched at the MS1 level (output from `precursor_matching.R`).
- `l6`: Matching results *before denoising*.
- `tolerance`: MS/MS fragment ion tolerance in Daltons (default: `0.05`).

---

##### **What the Function Does**

- Iterates over each matched feature and retrieves its associated annotated MS/MS spectra.
- Applies denoising at multiple frequency thresholds (`0.00` to `1.00` in steps of `0.01`) to generate subspectra.
- Stores these thresholded spectra in the folder: `Denoised_spectra_for_tuning/`.

- For each feature:
  - Matches all thresholded spectra against the best-matching reference spectrum (identified before denoising).
  - Computes similarity scores using dot product-based metrics and stores results in `After_denoising_matches/`.

- Calculates:
  - **Signal reduction**: Loss in matching fragments post-denoising.
  - **Noise reduction**: Reduction in unmatched fragments (interpreted as noise).
  - **FMR (Fragment Match Rate)**: Ratio of matching to total fragments before and after denoising.

- Applies **Pareto front analysis** to identify optimal trade-offs between signal retention and noise reduction.
  - If applicable, also uses **knee-point detection** and **differential evolution optimization (DEoptim)** as backup strategies.

- Stores results as:
  - CSV files of all evaluated matches (`pareto_results/csv/`)
  - Pareto front plots for each feature (`pareto_results/pdf/`)
  - Final optimal scores in a merged summary table.

- Calculates:
  - The percentage increase in similarity scores (after vs. before denoising)
  - Filters features with improved scores and positive signal retention for downstream evaluation.

---

##### **Output**

Returns a filtered data frame containing:
- Optimal frequency threshold (`freq`)
- Matching score before and after denoising
- Signal and noise reduction metrics
- Fragment match rates (FMR)
- Similarity scores
- Final selected features based on Pareto optimization

---


```{r tuning_module}
l7 = tuning_module(folder_path, l4, l5, l6, 0.05)
```

### Step 8: How stable are the derived recurrence frequencies? `sensitivity_analysis()`

This function performs a **sensitivity analysis** to evaluate how different fragment frequency thresholds impact spectral matching quality. It compares matching metrics before and after denoising across recurrence frequencies and identifies the most stable range for parameter selection.

---

#### **Inputs**

- `l7`: A data frame containing matching scores and metadata before and after denoising (output of `tuning_module()`).
- `folder_path`: Directory where the sensitivity analysis output files will be saved.

---

#### **What the Function Does**

- Filters the set of features that show **improvement in matching score** after denoising. If none show improvement, it uses the entire feature set.
- Loads denoised match results (`After_denoising_matches/`) for each selected feature and appends matching scores and metadata.
- Computes **scaling factors** to account for the frequency distribution of features, and applies these to adjust:
  - `Matching_Score` (scaled)
  - `Noise fragments` (scaled)
- For each recurrence frequency, calculates:
  - **Median scaled matching score**
  - **Median number of noise fragments**
  - **Percent noise reduction** relative to baseline (frequency = 0)
- Performs **Wilcoxon rank-sum tests** to assess significance of metric changes at each frequency compared to the optimal frequency.
- Identifies a **stable frequency range** where:
  - Metric performance remains optimal
  - Feature count is sufficiently high (≥ 80% of max observed)
  - No significant drop in performance (based on p-values)

---

#### **Output**

- Two summary plots:
  1. **Similarity Score Sensitivity**  
     Shows how median matching score varies across frequencies with the stable range highlighted.
  2. **Noise Fragment Sensitivity**  
     Shows changes in noise fragments with statistically significant reductions marked with `*`.

- These plots are saved as:
  - `Sensitivity_Analysis_Matching_Score.pdf`
  - `Sensitivity_Analysis_Noise_Fragments.pdf`

- **Output Location:**  
  Both plots are saved in the specified `folder_path`.

---

```{r sensitivity-plots, fig.width=10, fig.height=8, message=FALSE, warning=FALSE}
l8 <- sensitivity_analysis(l7, folder_path = folder_path)
```

> **Note:**  
> Owing to the **small number of features** selected for this vignette, the sensitivity curves presented here may not be highly informative. Specifically, the `l7` dataset contains **only one feature** that showed an improvement in matching score after denoising. As a result, this **single datapoint** was used for further analysis.
> 
> For more representative sensitivity plots generated using full experimental datasets (as used in the manuscript), please refer to the examples available at the following Google Drive link:  
> [**Sensitivity Curve Examples (Google Drive)**](https://drive.google.com/file/d/1eDWAx6eBN5bjyD_8ZQ3jTn4J78b_cT3K/view?usp=sharing)

Users can now proceed to the next step using either of the following:

- The **optimal frequency threshold** identified from the summary statistics in **Step 7: Tuning Module**  
  (which was `0.29` in our example), **or**
- A **custom frequency threshold** selected based on the sensitivity analysis results.

This threshold will be used to denoise the experimental spectra before the final round of matching and downstream analysis.

### Step 9: Generating denoised spectra and storing them as mzML/mzXML files `generate_denoised_spectra()`

This function performs **frequency-based denoising** of MS/MS spectra and exports the filtered spectra into `.txt` and `.mzML` formats for downstream processing.

---

#### **Inputs**

- `aggregate_list`: Output from `label_individual_spectrum()` — a list of replicate MS/MS spectra with fragment-level frequency annotations.
- `folder_path`: Path to the working directory where denoised spectra will be saved.
- `custom_threshold`: *(Optional)* Frequency threshold to identify signal fragments (default = `0.1`).  
  Only fragments with frequency **≥ `custom_threshold`** will be retained in the denoised spectra.
- `ion_mode`: Ionization mode; `"pos"` for **positive mode** *(default)* or `"neg"` for negative mode.

---

#### **What the Function Does**

- Iterates through each feature and its replicate MS/MS spectra.
- Applies a user-defined frequency threshold (default = `0.1`) to retain **signal fragments**.
- For each replicate scan:
  - Filters and sorts the retained fragments by *m/z*.
  - Saves the cleaned spectrum to a `.txt` file in:
    ```
    <folder_path>/Denoised_spectra_txt/<Feature_ID>/<Feature_Scan>.txt
    ```

- Then for each **sample**:
  - Combines the denoised replicate spectra into a single `Spectra` object.
  - Exports the sample-level spectrum to an `.mzML` file using the `MsBackendMzR` backend.
  - Files are saved in:
    ```
    <folder_path>/Denoised_spectra_mzML/
    ```

---

#### **Output**

- `.txt` files for each replicate spectrum.
- `.mzML` files for each sample combining all denoised scans.
- Output folder structure:
  - `Denoised_spectra_txt/`
  - `Denoised_spectra_mzML/`

---

#### **Example Usage**

```r
generate_denoised_spectra(
  aggregate_list = l3,
  folder_path = "path/to/project/",
  custom_threshold = 0.1,
  ion_mode = "pos"
)
```

> **Tip:** You may change `custom_threshold` based on the optimal value obtained from tuning or sensitivity analysis (e.g., `0.29`). Again this step requires sps_top80_tic_2 and step 2 must be run before this step. 

```{r last step}
l9 = generate_denoised_spectra(l4, folder_path, 0.29, ion_mode = "neg")
```

With the execution of `generate_denoised_spectra()`, the denoising workflow is complete. The output includes both `.txt` and `.mzML` formats of the filtered MS/MS spectra, organized by feature and sample. These denoised spectra are now ready for downstream tasks such as spectral library matching, biomarker discovery, or machine learning-based classification. By leveraging fragment recurrence frequencies across replicates, this approach helps retain high-confidence signal peaks while effectively minimizing noise, thereby improving the interpretability and quality of MS/MS-based metabolomic analyses.

