---
title: "Demonstration of the tuning module of the package"
author: "Shayantan Banerjee"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Demonstration of the tuning module of the package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
options(repos = c(CRAN = "https://cloud.r-project.org/"))

# Suppress title check if necessary
options(rmarkdown.html_vignette.check_title = FALSE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


## Test Dataset for tuning

The test dataset used in this vignette is hosted at [https://zenodo.org/records/15132924](https://zenodo.org/records/15132924). Users need to download the file and unzip the folder contents on their local system.

- The `test_data_MTBLS606` folder contains a folder named `mzml_files/` and two files: `Sample-and-feature-wise-RT-tolerance.txt` and `feature_list.txt`. The `feature_list.txt` file has three columns: "ID", "mz", and "RT". The additional file `Sample-and-feature-wise-RT-tolerance.txt` contains the sample-wise RT tolerances for all features. The number of rows in both files is the same, while the `Sample-and-feature-wise-RT-tolerance.txt` file has twice as many columns as `feature_list.txt`, with `RT_min_` and `RT_max_` appended to every sample name.
  
## Install the dependencies
A simple and automated way to install and load all the dependencies is to run the following function
```{r initial-setup, eval=FALSE}
install_and_load_dures_dependencies()
```
This should be followed by installing the actual package using devtools as shown below

```{r initial-setup-contd, eval=TRUE}
if (!requireNamespace("devtools", quietly = TRUE)) {
  install.packages("devtools", quiet = TRUE)
}
devtools::install_github("BiosystemEngineeringLab-IITB/dures", auth_token = NULL)
library(dures)
```

## First Test Example
## Test Dataset

The test dataset can be downloaded from the data deposited at [https://zenodo.org/records/15132924](https://zenodo.org/records/15132924) (folder name: `test_data_MTBLS606/`).

### Workflow Overview

1. **Reading mzML Files**  
   The first step of the workflow involves reading mzML files and extracting MS/MS spectra within the specified *m/z* and *RT* tolerance.

2. **Extracting and Concatenating Spectra**  
   The extracted spectra are then concatenated for all replicate spectra belonging to a given feature.

3. **Filtering Features**  
   Features that do not have any MS/MS spectra extracted, based on the specified *RT* and *m/z* tolerance, are removed from the list.

```{r, execute first step of the workflow}
# Define the path to the  folder (replace with the actual path on your system)
folder_path <- "/data1/sbanerjee/test_data_MTBLS606/"

# Check if the directory exists
if (!dir.exists(folder_path)) {
  stop("The directory test_1 does not exist. Please modify the path to the actual location of the test_1 folder")
}

l1 <- preprocess(folder_path = folder_path, tol_mz = 25, tol_rt = 0.1)
```


More details on what the list `l1` contain, is provided in the other vignette titled: "Demonstration of the package dures"

```{r, execute intermediate step of workflow}

print(paste0("Length of stats file after removing features from which MS2 spectra couldn't be extracted: ", as.character(dim(l1$stats_file_ms2_only)[1])))
```
For purposes of this vignette, we will consider a random subset of 50 features from this list of 141 MS2 features. The following step can be skipped while runnning the code on an actual experimental dataset.

```{r, Subsetting the feature list to reduce computation}
# Set seed for reproducibility
set.seed(123)

# Step 1: Get all IDs from the dataframe
all_ids <- l1$stats_file_ms2_only$ID

# Step 2: Sample 50 IDs
selected_ids <- sample(all_ids, 50)

# Step 3: Subset dataframe and reorder by selected_ids
subset_stats <- l1$stats_file_ms2_only[l1$stats_file_ms2_only$ID %in% selected_ids, ]
subset_stats <- subset_stats[match(selected_ids, subset_stats$ID), ]

# Step 4: Subset and reorder spectra list to match the same order
subset_spectra <- l1$spectra_ms2_only[selected_ids]  # list is named by ID

# Step 5: Store both in a new list
l1_subset <- list(
  stats_file_ms2_only = subset_stats,
  spectra_ms2_only = subset_spectra
)
```

From now on, we will use `l1_subset` for all subsequent analysis. The user may stick to `l1` to continue with the original set of MS2 features

### Step 2: Extracting Top x% TIC Spectra and Grouping Fragments

In the second step, we will use `l1` from the previous step as input to:

1. **Extract the top x% TIC spectra**, and
2. **Group fragments** within a specified **mass tolerance**.

The user can modify the following parameters:
- **Top x% TIC value**: Default is **80%** (i.e., **0.8**).
- **Mass tolerance**: Default is **0.05 Da**.

---


```{r second step}
l2 = extract_raw_spectra(folder_path = folder_path, l1_subset, 0.05, 0.8)
```

The following code checks to ensure that the object `sps_top80_tic_2` is available in the local environment. This is not required while running the package

```{r check_and_assign, echo=TRUE, error=TRUE}
# Check if 'sps_top80_tic_2' exists before assigning
if (!exists("sps_top80_tic_2", envir = .dures_env)) {
  stop("Object 'sps_top80_tic_2' not found in .dures_env.")
} else {
  assign("sps_top80_tic_2", get("sps_top80_tic_2", envir = .dures_env), envir = .dures_env)
}

```

Please refer to the vignette titled "Demonstration of the package dures" to know more about the structure of `l2`

### Step 3: Consensus Spectrum Generation

In the **third step**, a **consensus spectrum** is generated using the **top 80% TIC spectra**, and the corresponding **fragment frequencies** are calculated.

- The **consensus spectrum**, also known as the **inter-spectrum aggregate**, is built using the same principle as the **intra-spectrum aggregate**.
- The output from this step includes:
  1. A **single spectrum per feature**.
  2. The **peak frequencies** for each feature.

```{r third step - most expensive}
l3 = call_aggregate(l2$sps_top_tic_2, 0.05, folder_path)
```
### Step 4: Labelling individual spectra for every feature

Using the **recurrence frequencies** calculated in **Step 3**, we will now label the fragments of every **top TIC spectra** for a given feature.

The output from this step is:

- A **list** containing all spectra for a given feature.
- The corresponding **fragment frequencies** for each labeled fragment.

This labeling process provides insights into how often each fragment appears across all spectra, allowing for better analysis of fragment patterns within the **top TIC** data. This can be an expensive operation, especially if there are a lor of fragments per feature.


```{r label_spec}
l4 = label_individual_spectrum(l3, folder_path, 0.05)
```
